---
title: "FoRTE baseline"
author:
  - Alexey N. Shiklomanov
  - Ben Bond-Lamberty
  - Jeff Atkins
  - Christopher M. Gough
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
    pdf_document:
      latex_engine: xelatex
bibliography: references.bib
csl: "../templates/ecology.csl"
abstract: |
  Forests in the US Upper Midwest have been a net C sink for a century, driven by their fast-growing, early successional species. 
  However, the composition of these regrown forests has started to shift toward mid-successional species. 
  Combined with changing climate and disturbance, this means that future biogeochemical and ecological dynamics in these forests are highly uncertain. 
  Demographic vegetation models provide a tool for analyzing this uncertainty and its implications. 
  Here, we evaluate the relative importance of model "structural" (which processes are represented, and how) and "parameter" (parameters used in those equations) uncertainty, asking: 
  (1) Which processes are most important to accurately model community succession and C cycling in Upper Midwest forests?
  (2) What are the relative contributions of parameter vs. structural uncertainty?
  We ran ensembles of the Ecosystem Demography model v2.2 (sampling over parameter uncertainty) from 1900 to 2000 with different representations of key processes important to competition for light.
  We then compared the magnitude of structural (across representations) and parameter (within each representation) uncertainty, and assessed which submodel-parameter combinations reproduced observed C fluxes and succession.
  On average, our simulations underestimated both net primary productivity (NPP) and leaf area index (LAI) after 100 simulation years, though NPP estimates at peak productivity (20-50 years into the simulations) were closer to observations.
  However, parameter uncertainty was large; at any given timestep, both NPP and LAI ranged from ~0 to >200% of their mean value.
  Although many simulations matched either the observed NPP or observed LAI at some point during the simulation, only a small minority of individual runs matched both.
  Structural uncertainty was less important than parameter uncertainty, but model configuration did influence the ecological dynamics within a given parameter set, and affected which parameters were most influential.
  We conclude that: (1) targeted data collection to constrain vegetation model parameters should be a higher priority than model development; (2) parameter sensitivity analyses are not necessarily generalizable across model structures; and (3) robust model validation requires scrutiny of multiple different variables, including those related to vegetation productivity, structure, and composition.
keywords: |
  Dynamic vegetation modeling; carbon cycle; forest succession
highlights: |
  These are the highlights.
---

```{r setup, include = FALSE}
library(knitr)
library(here)

fig <- function(...) {
  knitr::include_graphics(here::here("analysis", "figures", ...))
}

opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  out.width = "\\linewidth"
  ## fig.path = "../figures/"
)

## library(fortebaseline) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
n <- 0
mdcap <- function(text) {
  n <<- n + 1
  out <- sprintf("**Figure %d**: %s", n, text)
  if (grepl("gfm", knitr::opts_knit$get("rmarkdown.pandoc.to"), ignore.case = TRUE)) {
    out
  } else {
    invisible(NULL)
  }
}
m <- 0
htmlfig <- function() {
  m <<- m + 1
  out <- paste0("**Figure ", m, "**:")
  if (grepl("html", knitr::opts_knit$get("rmarkdown.pandoc.to"), ignore.case = TRUE)) {
    out
  } else {
    invisible(NULL)
  }
}
```

# Introduction

**Forests are a defining feature of the US North**.
This region is the most heavily forested region in the US, and these forests are an essential provider of employment and recreation opportunities and other ecosystem services for the region's people, and of habitat for the region's wildlife [@shividenko_2005_forest; @shifley_2012_forests].
In addition, the **forests of the Upper Midwest** in particular have been a **net carbon sink** for most of the past century, as fast-growing early successional species established and thrived following near-complete deforestation around the end of the 19th Century [@birdsey_2006_forest; @williams_2012_carbon].
However, the **fate of these forests is highly uncertain**.
Many Upper Midwest forests are undergoing a **shift in community composition from early-successional to mid-successional** species, and the ecological and biogeochemical consequences of this transition are not well-established [@gough_2016_disturbance].
In addition, these forests face a multitude of **direct and indirect pressures from both biotic and abiotic sources** [@shifley_2016_future], including longer and more severe droughts [@gustafson_2016_implications; @swanston_2017_vulnerability; @lienard_2016_us], non-native insects and pathogens [@lovett_2016_nonnative], and more. <!-- TODO More here? -->
Collectively, these processes have culminated in forests that are different from their pre-settlement climax counterparts [@stearns_2002_one; @wolter_2002_recent; @thompson_2013_four], and **our ability to make predictions about such non-analog environments based on the past is limited**.

**More reliable predictions can in theory be obtained by using dynamic vegetation models that explicitly represent processes involved in forest growth and mortality.**
Vegetation models fall broadly along the following spectrum of complexity, which is a direct consequence of trade-offs between **realism and process fidelity** on one hand and **tractability, computational demand, and data requirements** on the other [@hawkes_2000_woody]. <!-- TODO Might need a better ref here -->
On one side are relatively simple **"big leaf" models**---such as PNET, SiBCASA, and Biome-BGC---in which the vegetation at a particular location consists essentially of a single large "plant" whose characteristics are the (weighted) average of all the vegetation at that site <!-- TODO REFS -->
On the other end of the spectrum are **"individual-based" models** (a.k.a. "gap models"), which explicitly simulate multiple indivduals competing for resources at a single site [@shugart_2015_computer]. <!-- TODO other refs? -->
Examples of such models are LANDIS [@mladenoff_2004_landis; @scheller_2007_design] and UVAFME. <!-- TODO ref --> <!-- TODO Other examples? -->
Because these models can explicitly represent inter-specific differences in plant productivity, resource allocation, and stress tolerance (among others), as well as the competition that emerges out of these differences, they may be better able to represent changes in ecosystem-scale processes, especially in no-analog conditions [@purves_2008_predictive].
Between these two extremes are models that use approximations to attempt to capture the emergent biogeochemical and ecological outcomes of interactions between individual plants without the need for explicitly simulating each individual [@purves_2008_predictive].
One specific example of this approach is the **Ecosystem Demography model** [ED2; @moorcroft_2001_method; @medvigy_2011_predicting; @longo_2019_ed1], which is the focus of this study.
<!-- BBL: Perhaps mention ED2's influence in ESMs and LSMs -->

**Regardless of vegetation model complexity, their projections are inherently uncertain.** 
This uncertainty comes from many different sources, which can broadly be classified into the following categories:
_Driver_ uncertainty refers to uncertainty in data about processes not represented by the model (e.g. weather and climate for vegetation models).
_Initial condition_ uncertainty arises from the fact that models have to start somewhere, and the exact conditions at the place and time simulations begin are frequently unknown.
_Process_ or _structural_ uncertainty arises because vegetation models necessarily only represent a subset of all processes involved in plant biology, and that the process that are included can usually be represented in multiple different ways (i.e. which processes are included, and how they are represented).
Finally, _parameter_ uncertainty arises because of natural variability and imperfect calibration of the above process representations.
**The relative importance of these sources of uncertainty in ecological forecasts is a key question because it is directly related to future research priorities**.

**Many ecosystem modeling studies have looked at these categories of uncertainty independently.**
<!-- TODO Consider cutting 50-75% of these next few sentences. Probably way too much detail. -->
@dietze_2014_quantitative investigated which parameters were most important to ED2 predictions of C sequestration across a diverse range of sites in North America.
The found that the parameters contributing the most to overall predictive uncertainty were those related to growth respiration, mortality, stomatal conductance, and water uptake, and that parameter uncertainty varied with biome.
More recently, @raczka_2018_what evaluated parameter uncertainty over 100 year time scale at the Willow Creek Ameriflux site, and found that the most important parameters to model uncertainty were quantum efficiency of photosynthesis, leaf respiration, and soil-plant water transfer.
Another study looked specifically at contributions to uncertainty from parameters related to canopy radiative transfer, and found that parameters related to both canopy structure and leaf optical properties had a large impact on predictions of net primary productivity [@viskari_2019_influence].
{**Driver uncertainty** -- a few examples?}

**Other studies have investigated structural uncertainty.**.
Numerous vegetation model intercomparison studies have demonstrated that different models produce significantly different projections of overall land carbon sequestration [@friedlingstein_2006_climate; @friedlingstein_2014_uncertainties], response to CO~2~ fertilization [@zaehle_2014_evaluation; @medlyn_2015_using; @walker_2015_predicting], and soil C sequestration [@sulman_2018_multiple], among others.
In particular, @lovenduski_2017_reducing found that model structural uncertainty was the primary source of uncertainty in Earth system model projections of the land C sink.
This structural uncertainty has been attributed to differences in model representations of key processes, including canopy radiative transfer [@fisher_2017_vegetation], soil biogeochemistry [@wieder_2017_carbon; @sulman_2018_multiple], stomatal conductance [@franks_2018_comparing], and photosynthesis [@rogers_2016_roadmap].
**However, the extent to which specific processes contribute to model uncertainty is difficult to evaluate from model intercomparisons because different models are different from each other in too many different ways, meaning that structural effects are confounded with other uncertainty sources.**
**Moreover, comparative analysis of contributions of different types of uncertainty to model projections are rare in the ecosystem modeling literature**.
<!-- 
TODO Do we have any examples? Or are these studies entirely non-existent?

Ben's notes:which of the last two categories--structural versus parameter uncertainty--is more important (Fisher et al. 2015, https://onlinelibrary.wiley.com/doi/full/10.1111/j.1523-1739.2003.00614.x, Ricciuto et al. 2008).
-->

<!-- TODO This paragraph is janky as hell. Needs heavy revision. -->
(Chris/Jeff: Help)
Previous work found that an earlier version of ED had difficulty sustaining multiple plant functional types in long simulations [@fisher_2010_assessing].
One of the primary drivers of successional change in temperate forests is competition for light (REF Bazzaz 19XX) <!-- TODO REF -- Bazaaz? -->, and differences in plants' strategies for light acquisition and utilization are a key driver of niche differentiation in these ecosystems <!-- TODO REF -->.
Plants compete for light in three different ways:
Growing quickly (early successional, pioneer species),
structural complexity / morphological change / filling in gaps in existing canopies,
and adapting traits to maximize light (and other resource) capture and utilization.
To this end, we focused our structural uncertainty analysis on three processes related to competition for light:
Representation of horizontal competition;
representation of canopy radiative transfer (especially of diffuse radiation);
and trait plasticity with light level.

**This study focuses on the interaction between parameter and structural uncertainty in the ED2 model.**
Our study is organized around the following guiding questions:
(1) Which processes related to light utilization are most important to accurately modeling community succession and C cycle dynamics in temperate forests of the Upper Midwest?
(2) What is the cost of considering these processes, in terms of additional parameter uncertainty?
(3) What are the relative contributions of parameter uncertainty (data limitation) vs. structural uncertainty (theoretical limitation?)?
We ran ED2 ensemble simulations with a factorial combination of submodels related to radiative transfer formulation, horizontal competition, and trait plasticity.
We then used sensitivity and variance decomposition analyses to evaluate the contribution of parameter uncertainty for each model configuration, exploring the ecological implications of the results in the context of ongoing climate change.

# Methods

## Site description

We performed this study at the University of Michigan Biological Station (UMBS; Ameriflux site US-UMd, 45.5625$^{\circ}$, -84.6975$^{\circ}$), located in Northern Lower Michigan, USA.
The area surrounding the research station is 87% well-drained upland forest and 13% wetland [@bergen_2007_observing]; the focus of this study is on the former.
The landscape geography of the UMBS upland forest is 20.4% moraine, 37.8% high outwash plain, 31.3% low outwash plain, 5.7% lake-margin terrace, 3.6% ice-contact, and 1.2% lowland glacial lake [@bergen_2007_observing].
Most of the UMBS upland forest canopy is dominated by temperate deciduous early-successional species, most importantly _Populus grandidentata_ (bigtooth aspen) and, to a lesser extent, _Betula papyrifera_ (paper birch), with _Acer saccharum_ (sugar maple), _Acer rubrum_ (red maple), _Fagus grandifolia_ (American beech), _Tilia americana_ (basswood), _Betula alleghaniensis_ (yellow birch), _Fraxinus americana_ (white ash), _Tsuga canadensis_ (eastern hemlock), _Quercus rubra_ (northern red oak), _Pinus strobus_ (white pine), and _Pinus resinosa_ (red pine) existing in various fractions in the understory (and, in patches, in the canopy).
This composition is a legacy of the site's disturbance history:
The site was intensively logged in the late 1800s and early 1900s, and experienced regularly recurring fires until the mid-1920s, at which point a regime of active fire suppression started that has persisted to the present day. <!-- TODO ref -->
As a result, the average stand age in 2013 was 95 years. <!-- TODO ref -->
The majority of the forest that is aspen-dominated is undergoing succession to "northern hardwood" (maple, beech, basswood, birch, ash, hemlock), "upland conifer" (red and white pine), or "northern red-oak" ecotypes [@bergen_2007_observing].


## Ecosystem Demography Model (version 2.2)

For this study, we used the Ecosystem Demography Model, version 2.2 (ED-2.2).
A full description of the default configuration of the model is provided by @longo_2019_ed1.
Briefly, ED-2.2 solves the energy, water, and carbon cycles separately for each of multiple "cohorts" of trees of similar composition, size, and age sharing a micro-environment and disturbance history [@moorcroft_2001_method; @medvigy_2009_mechanistic; @longo_2019_ed1].
Besides the cohort-based approximation of gap dynamics, a distinctive feature of ED-2.2 is that the equations for energy, water, and carbon fluxes are defined in terms of total energy, water, and carbon, which ensures excellent conservation of mass in long-running (multidecadal) simulations [@longo_2019_ed1].
Various versions of the ED model have been evaluated in boreal [@medvigy_2011_predicting], temperate [@medvigy_2009_mechanistic; @dietze_2014_quantitative; @raczka_2018_what], and tropical biomes [@moorcroft_2001_method; @longo_2019_ed2], and have been applied in, among others,
paleoecological studies [@rollinson_2017_emergent],
free-air CO~2~ enrichment studies [@dekauwe_2013_forest; @miller_2015_alteration],
estimation of potential carbon stocks [@hurtt_2004_beyond_potential_vegetation],
and analyses of regional vegetation-climate feedbacks [@swann_2015_future].

The official ED2 source code is available at https://github.com/EDmodel/ED2, and the exact version used in this study (which includes minor revisions to accommodate the requirements of this study) can be obtained at https://github.com/ashiklom/ED2/tree/b048950971e91699b78fc566df133caf977fda32.

## Submodels

In this study, we ran a factorial combination of the following ED-2.2 configurations: (1) two-stream vs. multiple-scatter canopy radiative transfer models; (2) infinite vs. finite crown area (a.k.a. complete vs. partial shading); and (3) static vs. light-plastic traits.

Both canopy radiative transfer models in ED-2.2 resolve the full vertical radiation profile within a patch as a function of canopy structure (leaf and wood area indices, crown area, leaf angle distribution) and incident solar radiation, following the definitions in CLM 4.5 [@clm45_note].
Both models also use identical definitions for the optical properties of a single canopy layer as a function of leaf and wood optical properties, leaf angle distribution, canopy clumping, solar zenith angle, and leaf, wood, and crown area indices.
Both models represent direct (a.k.a. "beam") radiation as an exponentially decaying process, and solve for the diffuse (a.k.a. "hemispherical", "isotropic") radiation at each layer using a linear matrix equation.
Where the models differ is in the terms of this matrix equation.
The two-stream model (default) uses the two-stream approximation originally derived for atmospheric radiative transfer [@meador_1980_two] and later adapted to vegetation canopies [@dickinson_1983_land; @sellers_1985_canopy].
Meanwhile, the multiple-scatter model was derived from first principles by @zhao_2005_multiple specifically for vegetation canopies to address known biases and limitations of the two-stream approach [e.g. @wang_2003_comparison].
**Compared to the default two-stream model, the multiple-scatter model slightly increases diffuse light levels in the understory.**
Both of these models are described in detail in the Supplementary Information. <!-- TODO Exact reference to Appendix -->

The crown area submodel in ED-2.2 determines the nature of competition for light between cohorts.
In the default configuration ("infinite crown area", or "complete shading"), the leaf area of a cohort is distributed across the entire horizontal area of a patch.
This means that taller cohorts always receive more incoming radiation than shorter cohorts, even when the height difference is small.
This has been shown to excessively suppress competition from sub-dominant individuals and result in unrealistically homogeneous canopies [@fisher_2015_taking].
In the alternate configuration ("finite crown area", or "partial shading"), canopies take up only a fraction of the available horizontal area, which generally exposes the understory to more light.
The horizontal area of crowns is determined by allometric equations from @dietze_2008_changing.

The third submodel we evaluated was trait plasticity.
In the default configuration, all cohorts of a given plant functional type have the same parameters, regardless of environmental conditions.
This ignores the globally-documented intraspecific trait variability as a function of light level [@niinemets_2010_review; @keenan_2016_global]. <!-- TODO Also, classical ecological succession theory bazzaz_1979_physiological? -->
In the alternate configuration, as light level decreases (trees become more shaded), specific leaf area increases and V~cmax~ decreases, following empirical relationships from the tropics [@lloyd_2010_optimisation].

These sub-models are summarized in Table 1.

Table 1: ED-2.2 submodel descriptions

| Name                         | Description                                                                          | Color                     |
|------------------------------|--------------------------------------------------------------------------------------|---------------------------|
| **Crown model**              |                                                                                      |                           |
| `closed` (default)           | Cohort crowns take up entire patch area. Competition for light based only on height. | Light                     |
| `finite`                     | Cohort crown area is proportional to DBH according to PFT-specific allometry.        | Dark                      |
| **Radiative transfer model** |                                                                                      |                           |
| `two-stream` (default)       | Two-stream approximation [@sellers_1985_canopy; @clm45_note]                         | Primary (red, blue)       |
| `multi-scatter`              | Multiple-scatter approximation, following [@zhao_2005_multiple]                      | Secondary (green, orange) |
| **Trait plasticity**         |                                                                                      |                           |
| `static` (default)           | SLA and V~c,max~ are constant                                                        | Cool (blue, green)        |
| `plastic`                    | SLA increases, and V~c,max~ decreases, with light level                              | Warm (red, orange)        |


## Parameter descriptions

The full list of parameters in ED-2.2 is large, with over 100 parameters per plant functional type.
A full sensitivity and uncertainty analysis across all of these parameters is outside the scope of this study.
Instead, we selected a subset of parameters guided by (but expanding upon) previous ED2 sensitivity studies [@dietze_2014_quantitative; @raczka_2018_what; @viskari_2019_influence].

Three parameters are related to leaf-level physiology:
Following the enzyme-kinetic model of @farquhar_1980_biochemical, the rate of photosynthesis is the minimum of light-limited and enzyme-limited reactions.
The former are controlled by the _quantum efficiency_ parameter---maximum efficiency with which absorbed photosynthetically active radiation is converted to CO~2~.
The latter are controlled by _V~cmax~_, the maximum rate of carbon fixation by Rubisco.
The water demand of photosynthesis is controlled by the _stomatal slope_, the sensitivity of stomatal conductance of CO~2~ as a function of CO~2~ concentration and humidity at the leaf surface [@leuning_1995_critical; @ball_1987_model].
Three more parameters correspond to the respiration rates of _leaves_, _roots_, and "_growth maintenance_".

Two more parameters control carbon allocation:
One is the ratio of fine root to leaf biomass (`fineroot2leaf`, or `q`), and another is the ratio of "storage" carbon allocated to reproduction (`r_fract`).

Three parameters control various aspects of adult tree mortality.
The overall adult mortality rate in ED-2.2 ($M$) is the sum of density-independent mortality from aging ($M_I$, plants year^-1^), density-dependent mortality from C starvation ($M_D$), and mortality from cold/frost ($M_F$) (ED-2.2 technically also includes additional term for fire mortality, but we did not include fire in our simulations):

$$
M = M_I + M_D + M_F
$$

Density-independent mortality from aging ($M_I$) is a prescribed, PFT-specific parameter (`mort3`).
Density-dependent mortality from C starvation is calculated as a function of a cohort's C balance limitation:

$$
M_D = \frac{y_1}{1 + \exp\left[y_2 \left( \frac{C_k}{C_k^*} \right)\right]}
$$

where $C_k$ is the 12 month running mean C balance of the cohort, $C_k^*$ is the running mean ideal C balance if there was no light or water limitation, and $y_1$ (`mort1`; plants year^-1^) and $y_2$ (`mort2`; unitless) are PFT-specific parameters.
_Seedling mortality_ rate is prescribed as its own PFT-specific parameter.

Several parameters are related to canopy structure and radiative transfer.
_Specific leaf area_ (`SLA`) is used to convert leaf biomass to leaf area index, which in turn is used in a variety of calculations related to canopy radiative transfer and micrometeorology.
Canopy _clumping factor_ describes how evenly leaf area is distributed in horizontal space (1 being perfectly evenly; 0 being a "black hole" where all leaves are concentrated in a single point); and
leaf _orientation factor_ describes the average distribution of leaf angles (-1 being perfectly vertical, 1 being perfectly horizontal, and 0 being random).
Four parameters control leaf optical properties, namely the fractions of light reflected or transmitted in visible and near-infrared wavelengths (`leaf_(reflect|trans)_(vis|nir)`).
Details of how these parameters influence canopy radiative transfer are described in the Supplementary Information.

Other parameters:

- Fraction of litter carbon that enters the labile (fast) soil carbon pool (`f_labile`)
- Minimum plant height for reproduction (`repro_min_h`)
- Soil-plant water conductance, or "water availability factor" (`water_conductance`)
- C:N ratio in fine roots (`c2n_fineroot`) and leaves (`c2n_leaf`)
- Leaf turnover rate (evergreen-only; `leaf_turnover_rate`)
- Proportion of dispersal that is "global" (`nonlocal_dispersal`) <!-- TODO What is this? Can I drop it from the analysis? -->

The full list of parameters is shown in Table 2.

```{r param-table}
tbl <- readr::read_csv(here(
  "analysis", "data", "derived-data", "parameter-table.csv"
))
tbl <- subset(tbl, Ran == "yes", select = c(-unit_parsed, -Ran))
tbl[["ED Name"]] <- sprintf("`%s`", tbl[["ED Name"]])
kable(tbl, escape = FALSE, caption = paste0(
  "Table 2: Parameter descriptions. ",
  "For additional information, see the ",
  "[ED2 model wiki](https://github.com/EDmodel/ED2/wiki/PFT-parameters)."
))
```

## Plant functional types and parameterization

By default, ED-2.2 supports 17 different plant functional types, which divide plant species according to photosynthetic pathway (C3 vs. C4), growth form (grass vs. tree), leaf phenology habit (deciduous vs. evergreen), biome (e.g. temperate vs. tropical), and successional status (e.g. early, mid, late).
However, we limited our simulations to the four plant functional types that have any appreciable presence at UMBS: Early, mid, and late temperate deciduous trees and pines.
The species comprising these plant functional types are shown in Table 3.

Table 3: Plant functional type-species mappings

| Plant functional type | Species                 | Color  |
|-----------------------|-------------------------|--------|
| Early hardwood        | _Betula papyrifera_     | Violet |
|                       | _Populus grandidentata_ |        |
|                       | _Populus tremuloides_   |        |
| Mid hardwood          | _Quercus rubra_         | Blue   |
|                       | _Acer rubrum_           |        |
|                       | _Acer pensylvaticum_    |        |
| Late hardwood         | _Acer saccharum_        | Green  |
|                       | _Fagus grandifolia_     |        |
| Pine                  | _Pinus strobus_         | Yellow |

For each plant functional type, we generated a distribution of parameter values via the Predictive Ecosystem Analyzer (PEcAn) trait-meta analysis [@lebauer_2013_facilitating; see also @dietze_2014_quantitative; @raczka_2018_what].
Prior distributions for this meta-analysis were largely adapted from previous ED2 parameter uncertainty studies [@dietze_2014_quantitative; @raczka_2018_what; @viskari_2019_influence], and are shown in detail in Supplementary Information X. <!-- TODO Priors supplement; maybe describe entire meta-analysis? -->
Species trait data for this meta-analysis came from existing records in the BETY database [www.betydb.org, @lebauer_2017_betydb], as well as from publicly available records in the TRY database [www.try-db.org, @kattge_2011_try] and, specifically for leaf optical properties, from Shiklomanov (2019 {dissertation, chapter 3}).
All parameters not described in this section were set to their ED-2.2 PFT-specific default values.

## Modeling workflow

For each factorial combination of ED-2.2 configurations (described above), we ran 100 ensemble members from 1901 to 2000.
Each ensemble member was initialized from a "near-bare ground" condition: An equal number of seedlings of each plant functional type (see previous section) at the minimum resolvable size.
Driving meteorological data was 6-hourly CRU-NCEP <!-- TODO ref. More here? --> combined with an annual atmospheric CO~2~ record from Law-Dome ice core [@etheridge_1998_historical] and Mauna Loa observatory [@thoning_1989_atmospheric].
Soil texture was set to 92% sand, 7% silt, and 1% clay, per site-level observations in [@gough_2010_wood].
The initial soil moisture profile was set to the average soil moisture profile reported in the UMBS Ameriflux ancillary data (https://ameriflux.lbl.gov/sites/siteinfo/US-UMd).

<!-- TODO This would probably fit better somewhere else.  -->
Driver uncertainty is outside the scope of this study, and initial condition uncertainty is minimized by our experimental design, which is conditioned on a specific initial condition (near bare ground).

## Analysis of results

- Overall summary:
    - Overall parameter uncertainty -- Time-series plots of total annual GPP and NPP, and maximum LAI and AGB, averaged across ensemble members
    - Impact of parameter uncertainty on ecological trajectories -- plot of max annual LAI by PFT
    - Both compared against observations of recently observed LAI and NPP at UMBS (e.g. FASET, FoRTE)
- Uncertainty analysis:
    - Parameter vs. structural uncertainty -- Variance in mean annual GPP from 19XX to 19YY across ensembles within model structure, and compare against 
    - Breakdown of parameter uncertainty -- PEcAn sensitivity and uncertainty analysis [@lebauer_2013_facilitating; @dietze_2014_quantitative]
- Details of structural uncertainty:
    - Vertical profiles of radiation, traits?
    
All analyses for this work were performed using R 3.6 (see "Colophon" for full details).
Code and supporting data for reproducing this analysis are publicly available on the Open Science Framework (OSF) at https://osf.io/dznuf/.

# Results

## Impacts of structure alone

```{r , fig.cap = structure_default_cap}
structure_default_cap <- paste(
  htmlfig(),
  "Annual NPP and LAI from ED2 runs",
  "with default parameters",
  "across different configurations."
)
fig("structure-compare-default.png")
```

Different model structures agree for the first half of the simulation, at which point they started to diverge.
Post 1950, the combination of two-stream radiative transfer, finite canopy radius, and plastic traits had the highest sustained productivity, followed by the same configuration except with static traits, with the remaining configurations having very similar results.
Interestingly, the effects of changing individual model configurations are not additive --- for example, enabling trait plasticity had a different effect depending on whether the two-stream or multiple-scatter canopy RTM was used.

```{r , fig.cap = structure_default_light_cap}
structure_default_light_cap <- paste(
  htmlfig(),
  "Radiation profiles of cohorts from ED2 runs with",
  "default parameters, across different configurations.",
  "Each point represents a single cohort of the given PFT,",
  "position on the y-axis represents the height of that cohort,",
  "and position on the x-axis represents the relative light level",
  "received by that cohort."
)
fig("default-light-levels.png")
```

Differences in productivity across structures are perhaps best understood by looking at differences in resulting radiation profiles at different snapshots in time over the course of the simulation.
Enabling the finite canopy radius increased the availability of light for understory cohorts throughout the simulation, with the most notable effects up to 1950.
Under a closed canopy, switching from a two-stream to a multiple-scatter canopy radiative transfer model also enhanced the understory light environment under a closed canopy, but to a lesser extent, and the brightest understory occurred under the combination of the two-stream radiative transfer and finite canopy radius.
However, none of the configurations were able to sustain any PFTs other than early hardwoods by 1980, and only the unique combination of finite canopy radius and two-stream radiative transfer were able to sustain more than a single dominant cohort (2 early hardwood cohorts with static traits, and 3 with plastic traits).

## Default vs. median parameters

```{r param_dist, fig.cap = param_dist_cap}
param_dist_cap <- paste(
  htmlfig(),
  "Input parameter distributions from PEcAn trait meta-analysis."
)
fig("param-dist.png")
```

Parameter estimates from the trait meta-analysis were often substantially different from the ED2 defaults.
In many cases, median parameter estimates were different entirely because of the choice of prior distributions (though, in almost all cases, the distributions provided reasonably high likelihood to ED2 default parameter values).
However, in several important cases, parameter well-constrained by data produced values substantially different from the ED2 defaults.
For example, ED2 default values for SLA were too low for all PFTs except Early Hardwoods, and default values for Vcmax values were too low for all PFTs.

```{r default-median-flux, fig.cap = default_median_flux_cap}
default_median_flux_cap <- paste(
  htmlfig(),
  "Comparing results of ED2 with default parameters and",
  "posterior median parameters from trait meta-analysis",
  "across different strucures."
)
fig("default-median-fluxes.png")
```

```{r default-median-lai , fig.cap = default_median_lai_cap}
default_median_lai_cap <- paste(
  htmlfig(),
  "Comparing results of ED2 PFT-level LAI",
  "with default parameters",
  "posterior median parameters from trait meta-analysis",
  "across different strucures."
)
fig("default-median-lai.png")
```

The above differences in median parameter led to ED2 runs that differed from the default runs in several important ways.
First of all, throughout all simulations and across all model structures, autotrophic respiration was several times higher with median parameter estimates compared to default values.
This is driven primarily by higher median values for growth and leaf respiration across all PFTs (notably, the ED2 default is a growth respiration of _zero_ for all temperate hardwood PFTs).
In some cases, this is compensated by an increase in gross primary productivity (driven by higher Vcmax values across all PFTs), leading to similar net primary productivity values to the default parameter runs.
Second, the variation across model structures is much higher, with every change in model structure exerting a noticeable effect on flux dynamics.
The most dramatic example is that, with median parameter values, all of the simulations with the finite canopy radius produced much lower productivity and heterotrophic respiration than simulations with a closed canopy.
Third, even where net primary productivity is similar between default and median runs, the underlying forest composition is very different.
Namely, under the default parameterization, ED2 produces forests dominated by Early Hardwoods, with an understory of small mid- and even smaller late-hardwood trees that eventually die out.
However, with the median parameterization, ED2 starts out with forests dominated by _mid_-hardwood species that are quickly (pre-1910 to 1930, depending on configuration) replaced with pine, which remains the dominant canopy species throughout the rest of the simulation.
Exploratory analyses (results not shown / supplement) show that the main parameters responsible for this compositional trajectory are SLA (especially the higher SLA of pine and mid-hardwood), Vcmax, and growth respiration.
While the uncertainty in growth respiration is large, we emphasize that SLA and Vcmax are well constrained by data for all PFTs considered in this analysis.

## Structure vs. parameter uncertainty

```{r summary-ts-plot, fig.cap = summary_ts_cap}
summary_ts_cap <- paste(
  htmlfig(),
  "ED2 plot-level predictions of",
  "net (NPP) primary productivity and",
  "total leaf area index (LAI),",
  "by model configuration.",
  "The dashed lines indicate the 10th and 90th quantiles of results.",
  "Black dot with error bars is the observed mean and min/max value",
   "(TODO REF)."
)
fig("summary-ts-plot.png")
```

<!-- - Structure x parameter uncertainty -->
<!--   - Parameter uncertainty is much larger than structural uncertainty -->
<!--   - Figure: Pairs of NPP, LAI, and PFT fraction, for each structure (first draw all, then pick relevant subsets) -->
<!--   - Q: How many runs end up with the right composition (EH-dominated) at peak (~1925), ~1950, and 1999? -->

<!-- TODO Should one of these be Early Hardwood (or other PFT?) LAI/biomass fraction -->
<!-- TODO Drop some of these variables? -->

Model projections of net primary productivity (NPP) and leaf area index (LAI) varied with both parameter value and model structure (Figure X).
All model configurations exhibited a similar trajectory of rapid growth in the first 15-20 years followed by a gradual decline to a lower, stable value around 1975 (with individual ensemble members sometimes undergoing sudden collapse in specific years).
The ensemble spread within any single configuration---i.e. the parameter uncertainty---was similar and large across all configurations; both NPP and LAI varied 5-10 fold at any given timestep depending on parameter values.
Within all configurations, most (>90%) ensemble members under-predicted observed present-day NPP and LAI by year 2000, though individual ensemble members from some configurations over-predicted one or both variables.
Predictions of peak NPP (between 1910 and 1950 for most simulations) were closer to the observed present-day NPP (though also with large ensemble spread), but predictions of LAI during this period were still too low.

```{r pft-jitterplot, fig.cap = pft_jitterplot_cap}
pft_jitterplot_cap <- paste(
  htmlfig(),
  "Fraction of total aboveground biomass (AGB)",
  "of each plant functional type (PFT)",
  "predicted by different parameter-configuration combionations",
  "of ED2."
)
fig("pft-jitterplot.png")
```

```{r pft-ternplot, fig.cap = pft_tern_cap}
pft_tern_cap <- paste(
  htmlfig(),
  "Ternary plot of predicted PFT composition in 1920, 1950, and 1980.",
  "Ternary coordinates are as follows:",
  "L is Early Hardwood; T is the sum of Mid and Late Hardwood; R is Pine.",
  "Triangle fill color indicates the number of points within that bin."
)
fig("pft-tern.png")
```

Trajectories of species composition also varied widely by model configuration (Figure X, Y).
Under all configurations, most ensemble members were dominanted by a single PFT, and the sustained presence of more than 2 PFTs was limited to only a handful specific ensemble members across all configurations.
However, _which_ PFT was dominant varied depending on the parameter combination, and any of the four PFTs became the dominant one some subset of simulations.

```{r pairs-time, fig.cap = pairs_time_cap}
pairs_time_cap <- paste(
  htmlfig(),
  "ED2 predictions of annual mean NPP and growing-season LAI in 1920, 1950, and 1999",
  "by model type.",
  "Each point indicates a single ensemble run with a different set of parameters.",
  "Solid line is the best-fit generalized additive model (GAM).",
  "Dashed lines indicate the error bounds of present-day observed NPP and LAI.",
  "Text labels match the rows in Figure X below."
)
fig("npp-lai-pairs.png")
```

<!-- TODO Use a logarithmic, rather than a linear fit? -->

The emergent relationship between NPP and LAI was similar across model configurations early in the runs, but became more sensitive to model configuration later into the runs.
Specifically, compared to runs with the default closed canopy, runs with a finite canopy were more likely to have higher LAI values for the same level of NPP, especially later into the simulation.
Although many simulations matched _either_ the observed net primary productivity or the observed leaf area index, only a small minority of individual runs matched _both_.

```{r lai-pft-plot, fig.cap = lai_pft_cap}
lai_pft_cap <- paste(
  htmlfig(),
  "PFT-level predictions of leaf area index (LAI) by model configuration and parameterization.",
  "Each row of plots is a simulation with the same set of parameters across model configurations.",
  "Letters correspond to the labels in Figure 3.",
  "Note that these are predictions of total leaf area index summed across all cohorts of the same PFT,",
  "meaning that higher LAI in one PFT does not necessarily indicate dominance, size, etc."
)
fig("lai-pft.png")
```

Model structure and parameters interacted to determine the community assembly and dynamics of each simulation, with consequences for overall ecosystem productivity that varied by model structure (Figure X).
Parameter sets **A** and **B** both resulted in similar patterns of immediate and complete dominance by mid- and late-hardwood, respectively, followed by gradual decline to near-zero LAI.
In both of these simulations, peak productivity was higher and was sustained for longer when the canopy was closed rather than finite, and this difference was larger under a two-stream rather than a multiple-scatter radiative transfer model.
Trait plasticity had relatively little effect on most of these two simulations, though in parameter set **B**, under the combination of closed canopy, multiple-scatter radiative transfer, and plastic traits, pine overtook the dominant late hardwood PFT starting around 1975.
Both of these parameter sets also had the lowest LAI of the five selected here, with a peak LAI of 1 in **A** and 2 in **B**.
Parameter set **C** showed much more variety in compositional dynamics with model structure:
In all cases, mid-hardwoods began as the dominant species and were eventually overtaken by pine under a closed canopy and plastic traits, or early hardwood in all other cases (though with a closed canopy and static traits, the final LAI was low).
Parameter sets **D** and **E** had the highest productivity of these five sets, and were characterized by strong dominance of Early Hardwood and Pine, respectively.

## Uncertainty analysis

<!-- - Details of parameter uncertainty -->
<!--   - Figure: Parameter sensitivity and uncertainty analysis -->
<!--   - Some parameters are always important. Other parameters matter more for different structures. -->


```{r within-vs-across, fig.cap = within_vs_across_cap, fig.dim = c(9, 4), eval = FALSE}
within_vs_across_cap <- paste(
  htmlfig(),
  "Comparison of parametric uncertainty within ensembles (colored bars, colored by model type)",
  "and \"structural\" uncertainty (variance in ensemble means; black bar) by output variable.",
  "Output variables are expressed as growing season averages for all years after 1910."
)
readd(within_across_plot_knit)
```

<!--
Except for the configurations with persistently low productivity, parameter uncertainty (across ensembles within each model configuration) was much larger than structural uncertainty (variance in ensemble means across configurations; Figure 4).
The relative importance of parameter uncertainty varied with both model configuration and the variable in question.
Allowing SLA and V~cmax~ to vary with light level (warm colors, as opposed to static traits -- cool colors) slightly reduced variability in primary productivity, slightly increased variability in Shannon diversity, and had mixed effects on aboveground biomass and total leaf area.
Enabling the finite canopy radius (dark colors; compared to closed canopy -- light colors) dramatically decreased variability in primary productivity and Shannon diversity and increased variability in leaf area, and had mixed effects on aboveground biomass.
reduced variance in aboveground biomass, diversity, and primary productivity.
Finally, using a multiple-scatter (secondary colors) rather than a two-stream (primary colors) radiative transfer scheme increased variance in leaf area index, aboveground biomass, and diversity, but did not have a consistent effect on variance in productivity productivity.
-->

```{r partial-variance-npp, fig.cap = partial_variance_npp_cap}
partial_variance_npp_cap <- paste(
  htmlfig(),
  "Contributions of individual parameters to predicted uncertainty in NPP,",
  "by model type.",
  "Panels on the left are for the \"peak\" NPP period (averaged over 1925-1950),",
  "and on the right are for the end of the simulation (averaged over 1975-1999)",
  "Each panel shows the top 8 parameter values in terms of their variance contribution.",
  "Input parameter distributions are shown in Figure 1.",
  "Note that x-axis scales vary across panels."
)
fig("partial-variance-npp.png")
```

<!-- Model sensitivity to specific parameters ["elasticity" _sensu_ @lebauer_2013_facilitating] varied depending on the model configuration and the target output variable (Figure 5). -->
<!-- The parameter with the single highest elasticity to NPP and LAI across all model configurations was specific leaf area (SLA), though the PFT whose SLA was most important varied with model configuration. -->
<!-- Other frequently (but not always) important parameters were the leaf C:N ratio (`c2n_leaf`), ratio of fine root to leaf biomass (`fineroot2leaf`), leaf and root respiration rates, and leaf optical properties (`leaf_reflect_vis/nir`). -->
<!-- TODO Dig into model configurations. -->

The parameters that contributed most to overall model predictive uncertainty ("partial variance") varied with model structure and time period considered.
In general, uncertainties in pine growth respiration factor and early hardwood plant-soil water conductance contributed the most to parameter uncertainty in NPP during the "peak" period (1925-1950).
{TODO: Fill in the rest of this text...}

Parameter sensitivity for total LAI was similar to that of NPP (results not shown).

<!-- This is old text... -->
<!-- ...differed from parameters to which the model was most sensitive (Figure 5), largely driven by the fact that many of the parameters with the highest sensitivity (e.g. SLA, V~cmax~) were also well constrained by data (Figure 1). -->
<!-- Leaf optical properties of at least one PFT were among the top 10 most important parameters across all configurations, and in multiple cases, were the most important. -->
<!-- Other often important parameters were plant-soil water conductance (`water_conductance`), parameters related to adult mortality (`mort1,2,3`), and growth respiration. -->
<!-- Again, patterns in partial variance were largely idiosyncratic across model configurations. -->
<!-- TODO Dig into this more -->
<!-- TODO Check this analysis. Look at pairs plots of trait vs. variable, and my SA spline fits. -->

{TODO: Basic analysis of contributions from parameter interactions, a la Raczka et al. 2018? I expect these will be important.}

# Discussion

The overarching objective of this study was to investigate how model structure and parameters interact in multidecadal simulations of forest ecosystem productivity and community assembly.

## ED2.2 vs. reality

Our first question was: To what extent can the model, under any configuration, reproduce the observed reality?
**Ensemble mean trajectory matches traditional understanding of succession**---rapid growth to peak productivity, followed by gradual decline [@odum_1969_strategy_ecosystem_development]---but **happens too quickly** (Figure TODO);
real ecosystems sustain peak productivity for longer [@gough_2016_disturbance].
This idea that ED2.2's main issue is that it is moving too quickly is supported by the fact that ED2.2 generally underestimates productivity after 100 years, but model **predictions at peak productivity are close to reality**.
One reason ED2.2 is moving too quickly through succession is **missing mechanisms for sustaining productivity through early succession**.
In particular, recent evidence points to the **important role of low-severity, high-frequency disturbances** in sustaining ecosystem productivity [@gough_2016_disturbance].
ED2.2 has a mortality scheme, and indeed, parameters related to mortality were often among the top 10 most important parameters for predictive uncertainty (Figure 6).
<!-- Maybe move this elsewhere? -->
<!-- TODO Look at mortality across ED runs. Does it relate at all to results? -->
However, **our simulations did not include other sources of disturbance**, such as partial harvests, extreme weather, and damage from insects and pathogens, which are common to this region [@amiro_2010_ecosystem] and which may contribute to sustaining productivity and dominance of early-successional forests [@gough_2016_disturbance].
Future work will: (1) Calibrate ED2.2 disturbance response to experimental observations (FoRTE), and (2) run long-term simulations with prescribed disturbance frequencies and see how they affect ecological dynamics.

- Figure 3:
  - For the same LAI, closed canopy had higher productivity than finite. Why?
    - Not zero-sum; finite canopy hurts dominant trees but doesn't improve subdominant species that much <!-- TODO Need support for this! -->
  - Importance of using multiple observational constraints. Ecological equifinality is a real problem.

## Drivers of uncertainty

**Parameter uncertainty was the dominant driver of uncertainty in our analysis**.
Different parameter combinations led to **huge ranges in ensemble predictions of ensemble variables** -- NPP ranged from 0 to almost 2x the observed, and LAI was up to 3x observed (though more tightly clustered around zero) (Figure TODO).
Moreover, different parameter combinations produced **qualitatively different ecological narratives** (Figure TODO).
**Parameter uncertainty has implications beyond C cycling**---differences in species composition and relative abundance have implications for forest management and wildlife conservation, among others.

The dominance of parameter uncertainty over structural uncertainty in our results suggests that **constraining parameter uncertainty should be a higher priority than model development**.
Parameter uncertainty can be constrained in two different ways:
Using observational data if available;
and, for parameters "that cannot be directly observed", using "quantitative/algorithmic" model calibration techniques (e.g. Fer, other papers).
For example, leaf optical properties (i.e. reflectance, transmittance) are both measurable and highly influential, and therefore should be a priority for new data collection.
In contrast, "growth respiration factor" is an empirical coefficient and not measurable, but highly influential (c.f. Dietze, Raczka), and therefore should be constrained through calibration.

(
So **which parameters are a research priority**?
Leaf optical properties often do not get a lot of attention, but are important [see also, @viskari_2019_influence].
Also, soil-plant water conductance and growth respiration -- both of these were important in @dietze_2014_quantitative.
(Compare also to @raczka_2018_what, @pandit_2018_optimizing).
)

**Parameter uncertainty is not synonymous with sensitivity**.
Many of the parameters to which model outputs are most sensitive are already well constrained by data and do not contribute as much to predictive uncertainty (Figure TODO).
SLA and Vcmax fall into this category.
Although parameter uncertainty is still large, this means that **targeted data collection works and that we are moving in the right direction**!
This speaks to the value of looking at not just sensitivity, but partial variance [@lebauer_2013_facilitating; @dietze_2014_quantitative].

Although parameter uncertainty dominated, **structural and parameter uncertainty interacted in important ways**.
Individual model structure changes had relatively modest impacts.

* Interactive effects of different structures -- not additive
* Interactions between parameter and structural uncertainty -- 
  - For example, the relative importance of different parameters to predictive uncertainty (Figure TODO) differed across model structures (Figure 3).

Collectively, this means that **sensitivity and uncertainty analyses are not necessarily generalizable across models**.

# Conclusions

<!-- TODO -->

# Acknowledgements

<!-- TODO Fill in Acknowledgements -->
This project funded by NSF grant.
Cyberinfrastructure provided by Pacific Northwest National Laboratory (PNNL).
Data from University of Michigan Biological Station (UMBS).
Data from TRY (TODO: Specific TRY statement).

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository(here::here())
```

View this project on GitHub in repository [ashiklom/fortebaseline](https://github.com/ashiklom/fortebaseline)
(this manuscript file is in [`analysis/paper/paper.Rmd`](https://github.com/ashiklom/fortebaseline/blob/master/analysis/paper/paper.Rmd))
and in Open Science Framework (OSF) at https://osf.io/dznuf/ .
