---
title: "FoRTE baseline"
author:
  - Alexey N. Shiklomanov
  - Jeff Atkins
  - Ben Bond-Lamberty
  - Christopher M. Gough
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(fortebaseline) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

# Introduction

Here is a citation [@bond-lamberty_2015_moderate] and another citation [@clm45_note;@norman_1979_modelling].
Another citation [@bond-lamberty_2015_moderate].

# Methods

## Site description

- Secondary successional forest -- bigtooth aspen (_Populus grandidentata_), northern red oak (_Quercus rubra_), red maple (/Acer rubrum/), paper birch (_Betula papyrifera_), eastern white pine (_Pinus strobus_)
- Average age -- 95 years (2013)
- NEP in UMBS tower is 1.58 MG C ha-1 yr-1 (0.80-1.98) (1999-2006), but with landscape variation (REF?)
- Heavily logged in late 1800s and early 1900s, disturbed by fire until 1923
- Present day composition typical of upper Great Lakes region
- Previously, the site of the FASET experiment [@gough_2013_sustained; @gough_2008_multi]

## Model description: ED2

The Ecosystem Demography Model, version 2 (ED2) [@medvigy_2009_mechanistic].

Summary of common features (can move into supplement if necessary):

- Cohort-based scaling approach
- Photosynthesis / stomatal conductance
- Allometries

In this study, we run a factorial experiment of ED2 with the following features enabled or disabled:

- Radiative transfer
  - Two-stream model
    - Same general approach as in CLM 4.5 [@clm45_note], but rather than solving for just two layers (sunlit and shaded), solve for each canopy layer.
    - Set up the two-stream equations for each layer as a linear system and solve the entire set of equations numerically
    - Similar to approaches used in other models: JULES [@mercado_2007_improving], ORCHIDEE-CAN [@naudts_2015_vertically], CLM(SPA) [@bonan_2014_modeling]
  - Multiple scatter model **TODO**

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```
