---
title: "FoRTE baseline"
author:
  - Alexey N. Shiklomanov
  - Jeff Atkins
  - Ben Bond-Lamberty
  - Christopher M. Gough
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(fortebaseline) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

# Introduction

Here is a citation [@bond-lamberty_2015_moderate] and another citation [@clm45_note;@norman_1979_modelling].
Another citation [@bond-lamberty_2015_moderate].

# Background

```{r demo, eval = FALSE}
1 + 1
```

# Methods

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```
