---
title: |
  Structure and parameter uncertainty in centennial-scale projections
  of community-carbon cycling in a temperate forest
subtitle: Supplementary information
author:
  - Alexey N. Shiklomanov
  - Ben Bond-Lamberty
  - Jeff Atkins
  - Christopher M. Gough
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
header-includes:
  \renewcommand{\thetable}{S\arabic{table}}
  \renewcommand{\thefigure}{S\arabic{figure}}
---

```{r setup, include = FALSE}
fig <- function(...) {
  knitr::include_graphics(here::here("analysis", "figures", ...))
}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  out.width = "\\linewidth"
)
```

```{r umbs-map, fig.cap = umbs_map_cap}
### Figure S1
umbs_map_cap <- paste(
  "Location of University of Michigan Biological Station (UMBS; black dot)",
  "in the US Upper Midwest."
)
fig("umbs-map.png")
```

\clearpage

```{r structure-compare-light, fig.cap = structure_default_light_cap}
### Figure S2
structure_default_light_cap <- paste(
  "Radiation profiles of cohorts from ED2 runs with",
  "default parameters, across different configurations.",
  "Each point represents a single cohort of the given PFT,",
  "position on the y-axis represents the height of that cohort,",
  "and position on the x-axis represents the relative light level",
  "received by that cohort."
)
fig("default-light-levels.png")
```

\clearpage

```{r pft-histogram-plot, fig.cap = pft_histogram_cap}
### Figure S3
pft_histogram_cap <- paste(
  "ED-2.2 ensemble predictions of PFT aboveground biomass fractions."
)
fig("pft-histogram.png")
```

\clearpage

```{r fitobs-pie, fig.cap = fitobs_pie_cap}
### Figure S4
fitobs_pie_cap <- paste(
  "NPP predictions from best-fit parameters."
)

fig("pie-fitobs.png")
```

\clearpage

```{r params-fit-observed, fig.cap = params_fit_observed_cap}
### Figure S5
params_fit_observed_cap <- paste(
  "PFT parameters resulting in runs that fit observed NPP and LAI."
)

fig("params-fit-observed.png")
```

\clearpage

```{r diversity-pie, fig.cap = diversity_pie_cap}
### Figure S6
diversity_pie_cap <- paste(
  "NPP predictions from most diverse parameters."
)

fig("pie-diversity.png")
```

\clearpage

```{r params-high-diversity, fig.cap = params_high_diversity_cap}
### Figure S7
params_high_diversity_cap <- paste(
  "PFT parameters resulting in runs that produce NPFT > 2 and total LAI > 1."
)

fig("params-high-diversity.png")
```

\clearpage

Table: ED-2.2 default parameter values

| Parameter                | Unit                                  | Plant functional type     | Value                |
|--------------------------|---------------------------------------|---------------------------|----------------------|
| Specific leaf area (SLA) | m$^2$ kgC$^{-1}$                      | Early hardwood            | 14.4                 |
|                          |                                       | Mid hardwood              | 11.616               |
|                          |                                       | Late hardwood             | 28.8                 |
|                          |                                       | Pine                      | 2.88                 |
| V$_{c,\max}$             | $\mu$mol m$^{-2}$ s$^{-1}$            | Early hardwood            | 28.9                 |
|                          |                                       | Mid hardwood              | 24.7                 |
|                          |                                       | Late hardwood             | 9.90                 |
|                          |                                       | Pine                      | 16.1                 |
| C:N leaf                 | Unitless                              | Early hardwood            | 13.5                 |
|                          |                                       | Mid hardwood              | 19.3                 |
|                          |                                       | Late hardwood             | 18.1                 |
|                          |                                       | Pine                      | 78.0                 |
| Clumping factor          | Unitless (0 - 1)                      | Early, Mid, Late hardwood | 0.84                 |
|                          |                                       | Pine                      | 0.735                |
| Leaf orientation factor  | Unitless (-1 - 1)                     | All                       | 0                    |
| Refl. (VIS)              | Unitless (0 - 1)                      | All                       | 0.11                 |
| Refl. (NIR)              | Unitless (0 - 1)                      | All                       | 0.11                 |
| Trans. (VIS)             | Unitless (0 - 1)                      | All                       | 0.11                 |
| Trans. (NIR)             | Unitless (0 - 1)                      | All                       | 0.11                 |
| Quant. eff.              | mol CO$_2$ (mol photons)$^{-1}$       | All                       | 0.08                 |
| Stomatal slope           | Unitless                              | All                       | 7.20                 |
| Water cond.              | m$^2$ a$^{-1}$ (kgC root)$^{-1}$      | All                       | $1.9 \times 10^{-5}$ |
| Growth resp.             | Unitless (0 - 1)                      | Early, Mid, Late hardwood | 0                    |
|                          |                                       | Pine                      | 0.45                 |
| Leaf turnover            | year$^{-1}$                           | Pine                      | 0.333                |
| Root turnover            | year$^{-1}$                           | Early hardwood            | 5.77                 |
|                          |                                       | Mid hardwood              | 5.08                 |
|                          |                                       | Late hardwood             | 5.07                 |
|                          |                                       | Pine                      | 3.93                 |
| F. labile                | Unitless (0 - 1)                      | All                       | 0.79                 |
| Mort. C rate             | year$^{-1}$                           | All                       | 1                    |
| Mort. C bal.             | Unitless                              | All                       | 20                   |
| Mort. bg. rate.          | year$^{-1}$                           | Early hardwood            | 0.00614              |
|                          |                                       | Mid hardwood              | 0.00381              |
|                          |                                       | Late hardwood             | 0.00428              |
|                          |                                       | Pine                      | 0.00339              |
| Mort. seedling           | Unitless (0 - 1)                      | All                       | 0.95                 |
| Repro. C frac.           | Unitless                              | All                       | 0.3                  |
| Dispersal                | Unitless (0 - 1)                      | Early hardwood            | 1                    |
|                          |                                       | Mid hardwood              | 0.325                |
|                          |                                       | Late hardwood             | 0.074                |
|                          |                                       | Pine                      | 0.766                |
| Min. height              | m                                     | All                       | 18                   |
| Root:leaf                | Unitless                              | Early, Mid, Late hardwood | 1.13                 |
|                          |                                       | Pine                      | 0.346                |
| Leaf resp.               | $\mu$mol m$^2$ s$^{-1}$               | Early hardwood            | 0.419                |
|                          |                                       | Mid hardwood              | 0.359                |
|                          |                                       | Late hardwood             | 0.144                |
|                          |                                       | Pine                      | 0.233                |
| Root resp.               | $\mu$mol CO$_2$ (kg fine root)$^{-1}$ | All                       | 0.794                |

\clearpage

Table: Prior distributions for trait meta-analysis. Trait descriptions and units are given in Table 2 (main text).

| Parameter          | Plant Functional Type     | Distribution         |
|--------------------|---------------------------|----------------------|
| Specific leaf area | Early, Mid, Late hardwood | Gamma(5.13, 0.234)   |
| ''                 | Pine                      | Gamma(2, 0.2)        |
| Vcmax              | All                       | Weibull(1.7, 80)     |
| C:N leaf           | Early, Mid, Late hardwood | Gamma(15.53, 0.87)   |
| ''                 | Pine                      | Uniform(3, 300)      |
| Clumping           | All                       | Beta(3, 1.5)         |
| F. labile          | All                       | Beta(1.5, 1.5)       |
| Root:leaf          | Early, Mid, Late hardwood | LogNormal(0.21, 0.6) |
| ''                 | Pine                      | LogNormal(-0.25, 1)  |
| Growth resp.       | Early, Mid, Late hardwood | Beta(4.06, 7.2)      |
| ''                 | Pine                      | Beta(3, 3.6)         |
| Refl. (NIR)        | Early hardwood            | Beta(561.2, 776.7)   |
| ''                 | Mid hardwood              | Beta(34.8, 56.0)     |
|                    | Late hardwood             | Beta(393.2, 595.0)   |
|                    | Pine                      | Beta(35.0, 34.6)     |
| Refl. (VIS)        | Early hardwood            | Beta(258.9, 4463.9)  |
| ''                 | Mid hardwood              | Beta(10.1, 157.5)    |
|                    | Late hardwood             | Beta(72.4, 1282.6)   |
|                    | Pine                      | Beta(5.5, 53.0)      |
| Leaf resp.         | All                       | Gamma(1.5, 0.4)      |
| Trans. (NIR)       | Early hardwood            | Beta(173.2, 199.2)   |
| ''                 | Mid hardwood              | Beta(16.5, 17.3)     |
|                    | Late hardwood             | Beta(95.6, 85.6)     |
|                    | Pine                      | Beta(5.52, 13.8)     |
| Trans. (VIS)       | Early hardwood            | Beta(15.4, 561.1)    |
| ''                 | Mid hardwood              | Beta(2.1, 37.9)      |
|                    | Late hardwood             | Beta(8.8, 189.0)     |
|                    | Pine                      | Beta(1.1, 73.8)      |
| Leaf turnover      | Pine                      | Gamma(4, 8)          |
| Mort. C rate       | All                       | Gamma(1, 0.05)       |
| Mort. C bal.       | All                       | Gamma(1.47, 0.0578)  |
| Mort. bg. rate     | All                       | Uniform(0, 0.02)     |
| Dispersal          | All                       | Beta(1.1, 1.5)       |
| Leaf orient.       | All                       | Uniform(-0.5, 0.5)   |
| Quant. eff.        | Early, Mid hardwood, Pine | Weibull(3.32, 0.08)  |
| ''                 | Late hardwood             | Gamma(2, 22)         |
| Repro. C frac.     | All                       | Beta(2, 4)           |
| Min. height        | All                       | Gamma(1.5, 0.2)      |
| Root resp.         | All                       | Weibull(2, 10)       |
| Root turnover      | Early, Mid hardwood, Pine | Weibull(1.55, 1.5)   |
| ''                 | Late hardwood             | Weibull(1.55, 0.862) |
| Mort. seedlng      | All                       | Beta(3.61, 0.433)    |
| Stomatal slope     | All                       | LogNormal(2.3, 1)    |
| Water cond.        | All                       | LogNormal(2e-5, 3.5) |
